### 2021-07-20 paper/localpsf.tex
edits to rational positive definite section

>>>>

\subsection{Rational positive semi-definite modification}

In many problems of practical interest (e.g., Hessian approximation) $\mathcal{A}$ is symmetric positive semi-definite. However, $\widetilde{\mathbf{A}}$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable. Symmetry and positive semi-definiteness are important properties which should be preserved by the approximation, if possible. Also, a lack of symmetry and/or positive semi-definiteness in $\widetilde{\mathbf{A}}$ may prevent one from using popular and highly effective algorithms, such as the conjugate gradient method, to perform further useful operations involving $\widetilde{\mathbf{A}}$.
In this section we describe a method for modifying $\widetilde{\mathbf{A}}$ to make it symmetric positive semi-definite.

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the spectrum of $\widetilde{\mathbf{A}}_\text{sym}^+$ has certain desirable properties, which we will describe later in this section. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should typically be avoided, here it is fine to compute the inverse because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix, and there are fast, stable, scalable, and widely used algorithms for inverting H-matrices \cite{Hmatrixinverse}.

The rational matrix approximation \eqref{eq:rational_matrix_function} may be written as $\widetilde{\mathbf{A}}_\text{sym}^+ = f(\widetilde{\mathbf{A}}_\text{sym})$, where $f$ is the rational function 
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose the four unknown constants, $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of four equations:
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
f(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
\end{align}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $f$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\widetilde{\mathbf{A}}_\text{sym}$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}, a matrix-free Krylov method. This is fast because computing matrix-vector products with an H-matrix is fast, and because the eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ do not cluster near $\lambda_\text{min}$ or $\lambda_\text{max}$ (though the eigenvalues may cluster at zero).

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $f(0)=0$ is the unique minimum of $f$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $f$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $f(\widetilde{\mathbf{A}}_\text{sym}) = \widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. The equation $f(0)=0$ implies that the null space of $\widetilde{\mathbf{A}}_\text{sym}^+$ contains the null-space of $\widetilde{\mathbf{A}}_\text{sym}$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. The equation $f(\lambda_\text{max})=\lambda_\text{max}$ forces the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ to be close to the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$. I.e., $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. The equation $f(\lambda_\text{min})=|\lambda_\text{min}|$ ensures that the negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(\lambda_\text{min})$, then the positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ will be closer to the corresponding positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ (good), but the erroneous negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ will be amplified more in magnitude (bad). We find that $f(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

Using the definition of $f$ in \eqref{eq:rational_function_scalar}, and basic calculus and linear algebra, we may write Equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, and \eqref{eq:desired_property_4} in the following matrix form:
\begin{equation}
\label{eq:rational_3_linear_system}
\begin{bmatrix} 1 & 0 & 1/\mu \\ 0 & 1 & -1/\mu^2 \\ 1 & \lambda_\text{max} & 1/(\lambda_\text{max}+\mu) \end{bmatrix} \begin{bmatrix} c_0 \\ c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \\ \lambda_\text{max} \end{bmatrix},
\end{equation}
and we may write Equation \eqref{eq:desired_property_3} as follows: 
\begin{equation}
\label{eq:rational_extra_eq}
c_0 + c_1 \lambda_\text{min} + \frac{c_2}{\lambda_\text{min}+\mu} - |\lambda_\text{min}| = 0.
\end{equation}
Let $g(\mu)$ denote the left hand side of \eqref{eq:rational_extra_eq}, but with $c_0$, $c_1$, and $c_2$ now solving \eqref{eq:rational_3_linear_system} for the given value of $\mu$. I.e., within $g(\mu)$ the quantities $c_0$, $c_1$, and $c_2$ are no longer free variables; they now depend on $\mu$ implicitly. We solve the combined system, \eqref{eq:rational_3_linear_system} and $\eqref{eq:rational_extra_eq}$, by solving the one-dimensional nonlinear root finding problem
\begin{equation}
\label{eq:g_mu_zero}
g(\mu)=0
\end{equation}
with bound $\mu > |\lambda_\text{min}|$ and initial guess $\mu_0 = 2 |\lambda_\text{min}|$. Each evaluation of $g$ requires solving the $3 \times 3$ linear system \eqref{eq:rational_3_linear_system}, and once the solution $\mu$ is found, we solve \eqref{eq:rational_3_linear_system} one more time to get the final values of $c_0$, $c_1$, and $c_2$.
%
%Row 1 corresponds to the condition $f(M)=M$, row 2 corresponds to the condition $f(0)=0$, and row 3 corresponds to the condition $f'(0)=0$. To find the third row of $A$, we used the fact that the derivative of $f$ is given by $$f'(\lambda) = c_1 - \frac{c_2}{(\lambda+\mu)^2}.$$
%
%
%We want $f(\lambda)$ to have the following desirable properties:
%\begin{enumerate}
%	\item $f(\lambda)$ is nonnegative on $[m, M]$.
%	\item $f(\lambda) \approx \lambda$ when $\lambda$ is positive and not small.
%	\item $f(0)=0$ is the minimum of $f(\lambda)$ on $[m, M]$.
%	\item $f(\lambda) \le |\lambda|$ when $\lambda$ is negative.
%\end{enumerate}
%Property 1 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Property 2 ensures that $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. That is, the part of the spectrum of $\mathcal{A}$ that we expect to be well-approximated by the product-convolution approximation. Property 3 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ preserves the null-space (and near null-space) of $\widetilde{\mathbf{A}}_\text{sym}$. This is important because in practice, we often want to approximate operators $\mathcal{A}$ that have a large (possibly infinite) number of zero eigenvalues, and/or a sequence of eigenvalues that clusters at zero. Hessians in ill-posed inverse problems typically have both of these properties. Property 4 ensures that $f$ does not amplify erroneous negative eigenvalues $\widetilde{\mathbf{A}}_\text{sym}$. If $\mathcal{A}$ is positive semi-definite, then eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ can only be negative because of errors in the product-convolution approximation.
